Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations

### 2016-03-24 doc/manuscript.tex
update tex with new text

@@

\begin{figure}[htbp]
\centering
\includegraphics{../figs/fig_1.pdf}
\caption{\label{fig:simpletree}Relationship between axon structure and
extracellular potential. The top row shows simulated action potentials
travelling along a single axon with varying morphologies, as indicated
by the diagram in the left side of each subfigure. The lower row shows
responses of a population of 100 randomly branching and termnating
axons. Three sample morphologies are shown to the left of the lower row.
Action potential propagation direction indicated by the arrow. All
waveforms recorded at a distance of 150\(\mu\)m from the axon, and at
the depth indicated by plot position, spaced by 200\(\mu\)m.
(\textbf{A}) An action potential in an infinitely long, straight axon.
(\textbf{B}) Effect of a terminatiing axon. Action potential waveform
closest to the termination thickened for emphasis. (\textbf{C}) Effect
of a branching axon. The axon branches multiple times in the range of
200\(\mu\)m. Thicker waveform at the center of the bifurcation zone.
(\textbf{D}) Combined effect of bifurcations and terminations.
(\textbf{E}) Extracellular waveform in response a spike initiated
simultaneously in all axons. Spike initiation time indicated by the
raster lines at the top. (\textbf{F}) CSD calculated from the waveforms
shown in E. (\textbf{G}) Response to spontaneous activity superimposed
with a pulse of hightened spike rate, averaged over 100 trials. Example
spike times shown by the raster plot at top. (\textbf{H}) CSD calculated
from the pulse response in G.}
\end{figure}

@@

\emph{Mechanism underlying the observed properties I.} We begin by
looking at the response of a single axon tree to a single spike at
various locations along the tree, as shown in Figure 4. This can already
give us a rough idea about the origin of the observed polarity reversal.

The extracellular response at a recording location along a continuation
of the axon, i.e.~not close to a bifurcation or termination, is
triphasic (blue trace in Figure 4B). It has been noted that this shape
is approximated by the second derivative of the membrane voltage: the
first positive peak is the initial bend to the rising flank of the
action potential, the main dip corresponds to the high curvature at the
peak of the action potential, and the final, positive peak corresponds
to the bend from hyperpolarization to recovery.

The potential near a bifurcation (red trace in Figure 4B) has a more
biphasic shape. There is a small initial peak, but the response is
dominated by the second, negative and the third, positive peak. This can
be understood by comparing the shape to the previous example along a
non-bifurcating axon: The initial shape resembles the non-bifurcating
case, because the action potential is still mainly within the segments
before the bifurcation. As the action potential passes the bifurcation,
there are now two instances of the action potential, and the second and
third peak are doubled in size, leading to the more biphasic response.

The same reasoning can be applied to the case of the biphasic response
at a termination (green trace in Figure 4B): here there is an
approaching action potential, leading to an initial positive peak, but
as the termination is reached, the current flow ceases, and there is
only a reduced second peak, and the third peak, associated with the
action potential moving away from the recording location, disappears
completely.

A notable point here is that the potential at the bifurcation and
termination are similar in shape, but opposite in polarity. This
resembles the observed polarity reversal. It is thus not hard to imagine
that this effect is caused by the response being dominated by
bifurcations in the early part of NL where the axons branch out, and
then by terminations in the later part, where terminations are more
prevalent, with the two canceling out in the middle. We will test this
intuition in the following by examining the properties of a full bundle
of axons in a simplified analytical model.

\emph{Mechanism underlying the observed properties II.} In the case of a
simplified one-dimensional axon bundle, the EFP at a given location may
be described by the formula
\(\Phi(\mathbf{r},t) = \left[\left\{n(z)\cdot i_\lambda(z,t)\right\}\ast w(a,z)\right]_z\)
(see Methods and appendix). Assuming that the weighting function \(w\)
tends to 0 for large \(\left|z\right|\), and with the introduction of
the single axon current integral
\(I_\lambda(z,t)=\int_{-\infty}^zi_\infty(z',t)dz'\), the equation can
be split (for a derivation, see appendix) into two components:

@@

    \begin{itemize}
    \tightlist
    \item
      caption
    \item
      better alignment in D
    \end{itemize}
  \item
    Figure~\ref{fig:bigtree}

    \begin{itemize}
    \tightlist
    \item
      lower density of axons
    \item
      Add scalebar for time
    \item
      change color of axis label
    \item
      change color of box
    \item
      add PSTH
    \item
      align zeros in C
    \item
      add contour labels?
    \end{itemize}
  \item
    Figure~\ref{fig:distscaling}

@@

  \begin{itemize}
  \tightlist
  \item
    New Figure~\ref{fig:bigtree}

    \begin{itemize}
    \tightlist
    \item
      single axons

@@

      \begin{itemize}
      \tightlist
      \item
        multibranch
      \item
        endings and snips
      \end{itemize}
    \item
      Figures E and F
    \item
      PSTH
    \item
      CSD
    \end{itemize}
  \item
    Figure~\ref{fig:bigtree}

>>>>

\emph{Effects of axonal morphologies on extracellular action potentials}
In order to understand the way that common anatomical structures affect
the extracellular waveform associated with a single action potential, we
simulated simple axons that showed isolated instances of these features
in order to inspect their contributions to the EFP. We began by
simulating an infinitely long axon following a straight line path,
neither bifurcating nor terminating ({[}fig:simpletree{]}A).

Next we simulated the response of an axon that terminates
({[}fig:simpletree{]}B). Here the approaching action potential (top
traces) can be seen to have the same, triphasic shape as in the
non-branching case. When it approaches the termination, it gradually
deforms into a biphasic response, with the positive peak preceding the
negative peak.

The other structure commonly found in axons is a bifurcation. We
simulated a single axon which bifurcates three times on each branch
within a distance of 50 \(\mu\)m, leading to a total number of 8 fibers
leaving the bifurcation zone ({[}fig:simpletree{]}C). The potential near
this bifurcating axon also has a more biphasic shape. There is a small
initial peak, but the response is dominated by the second, negative and
the third, positive peak. This can be understood by comparing the shape
to the previous example along a non-bifurcating axon: The initial peak
resembles the non-bifurcating case, because the action potential is
still mainly within the segments before the bifurcation. As the action
potential passes the bifurcation, there are now several instances of the
action potential, and the second and third peak are multiplied in size,
leading to the more biphasic response. In the case of a single
bifurcation, this effect is also present, but the amplification of the
second and third peak relative to the first peak is not as notable as in
this example.

To understand how bifurcations and terminations interact when they are
present in the same axon (as they must always be in a real axon), we
simulated an axon with an identical number of bifurcations as in the
previous case, but then added terminations to all the fibers 200
\(\mu\)m after the bifurcation zone ({[}fig:simpletree{]}D). We found
that this leads to the same biphasic responses observed as observed in
the cases of the isolated anatomical features, with the triphasic
response occurring briefly in-between the two features.

A notable point here is that the potential at the bifurcation and
termination are similar in shape, but opposite in polarity.

\begin{itemize}
\tightlist
\item
  why this is interesting.
\item
  why we now move to a bundle
\end{itemize}

@@

\begin{figure}[htbp]
\centering
\includegraphics{../figs/fig_1.pdf}
\caption{\label{fig:simpletree}Relationship between axon structure and
extracellular potential. The top row shows simulated action potentials
travelling along a single axon with varying morphologies, as indicated
by the diagram in the left side of each subfigure. The lower row shows
responses of a population of 100 randomly branching and terminating
axons. Three sample morphologies are shown to the left of the lower row.
Action potential propagation direction indicated by the arrow. All
waveforms recorded at a distance of 150\(\mu\)m from the axon, and at
the depth indicated by plot position, spaced by 200\(\mu\)m.
(\textbf{A}) An action potential in an infinitely long, straight axon.
(\textbf{B}) Effect of a terminatiing axon. Action potential waveform
closest to the termination thickened for emphasis. (\textbf{C}) Effect
of a branching axon. The axon branches multiple times in the range of
200\(\mu\)m. Thicker waveform at the center of the bifurcation zone.
(\textbf{D}) Combined effect of bifurcations and terminations.
(\textbf{E}) Extracellular waveform in response a spike initiated
simultaneously in all axons. Spike initiation time indicated by the
raster lines at the top. (\textbf{F}) CSD calculated from the waveforms
shown in E. (\textbf{G}) Response to spontaneous activity superimposed
with a pulse of hightened spike rate, averaged over 100 trials. Example
spike times shown by the raster plot at top. (\textbf{H}) CSD calculated
from the pulse response in G.}
\end{figure}

@@

\emph{A simplified model of axon bundle field potentials} In the case of
a simplified one-dimensional axon bundle, the EFP at a given location
may be described by the formula
\(\Phi(\mathbf{r},t) = \left[\left\{n(z)\cdot i_\lambda(z,t)\right\}\ast w(a,z)\right]_z\)
(see Methods and appendix). Assuming that the weighting function \(w\)
tends to 0 for large \(\left|z\right|\), and with the introduction of
the single axon current integral
\(I_\lambda(z,t)=\int_{-\infty}^zi_\infty(z',t)dz'\), the equation can
be split (for a derivation, see appendix) into two components:

@@

    \begin{itemize}
    \tightlist
    \item
      add input intensity gaussian
    \item
      use labels E1, E2, E3, F1, F2
    \item
      switch E2 and E3
    \item
      more space between panels
    \item
      add scalebars to each panel
    \item
      superimpose CSD and EFP?
    \end{itemize}
  \item
    Figure~\ref{fig:bigtree}

    \begin{itemize}
    \tightlist
    \item
      lower density of axons
    \item
      Add scalebar for time
    \item
      change color of axis label
    \item
      change color of box
    \item
      add PSTH
    \item
      align zeros in C
    \item
      add contour labels?
    \end{itemize}
  \item
    Figure~\ref{fig:barnowl}

    \begin{itemize}
    \tightlist
    \item
      redo with new data
    \end{itemize}
  \item
    Figure~\ref{fig:distscaling}

@@

    \begin{itemize}
    \tightlist
    \item
      caption
    \item
      single axons

@@

      \begin{itemize}
      \tightlist
      \item
        multibranch
      \item
        endings and snips
      \end{itemize}
    \item
      Figures E and F
    \item
      PSTH
    \item
      CSD
    \item
      better alignment in D
    \end{itemize}
  \item
    Figure~\ref{fig:bigtree}
